---
title: "Introduction to the dplyr library"
author: "Dr. David Oury"
date: "March 31, 2017"
output:
  html_document:
    toc: yes
fontsize: 12pt
---

# Introduction 

Hadley Wickham, the author of `dplyr`, has provided a package which
contains the most common data manipulation functions
for working on data frames.
Briefly, these manipulations are ordering rows, selecting rows, 
selecting variables, modifying variables and summarizing variables.

The functions of the `dplyr` library not only manipulate data frames, 
but manipulate data tables (see library `data.table`) and 
databases, in particular some versions of SQL. 

Please experiment with these commands in an __R__ interpreter 
as you read about them. 
To begin, load the `dplyr` library. 
```{r, message=FALSE}
library(dplyr)
```

We demonstrate the main functions of the library using the `flights`
dataset of the `nycflights` library. 
```{r, message=FALSE}
library(nycflights13)
```

The structure of the dataset is displayed for reference. 
```{r}
str(flights)
```
Notice that the dataframe is of type `tbl_df`. 

The following sections correspond to each of the five data manipulation tasks: 
ordering rows, selecting rows, 
selecting variables, modifying variables and summarizing variables.

First though we describe the `tbl_df` function, which wraps a (local) dataframe
so that printing the dataframe never results in wrapped lines or thousands of 
printed rows. 
Recall that the `flights` dataframe is of type `tbl_df`. 

```{r}
class(flights)
```

This print statement omits some columns and only
displays ten (10) rows of the `flights` dataframe.
By default, when a dataframe of type `tbl_df` is printed 
only ten (10) rows are displayed and columns 
are omitted so that the the rows do not wrap. 

```{r}
print(flights)
```

If you want to print more rows use the `n` parameter to the `print` function. 

```{r}
print(flights, n=15)
```

The following `option` command can be used to restrict all printing: 

```{r}
options(dplyr.print_min=5)
flights
```

To display all columns use this `option` command:

```{r}
options(dplyr.width=Inf)
flights
```

Notice that the rows are wrapped and that only five (5) rows are printed. 
The value `Inf` can be replaced with any positive integer.
To reset this option so that rows are not wrapped use

```{r}
options(dplyr.width=getOption("width"))
```


# Ordering rows with the `arrange` function

The `arrange` function orders the rows of a data frame by the variables 
specified by the second and later arguments. 
For example, the following command returns the flight data frame sorted by 
the `month` variable. 
```{r}
arrange(flights, month)
```

By default, rows are sorted in _increasing_ order.
Use the `desc` function to sort in _decreasing_ order. 
```{r}
arrange(flights, desc(month))
```

Additional variables to sort by can be specified in the third and later arguments.
The second argument specifies the primary sorting variable. 
The third argument is used in case of ties, 
and so on for additional arguments. 
```{r}
arrange(flights, month, dep_delay)
```


# Selecting rows

Each of the functions 
`sample_n`, `sample_frac`, `slice` and `filter` 
retrieve a subset of the rows of a data frame. 

## Function `sample_n`

The `sample_n` function returns a random selection of records 
as determined by the `size` parameter.
The command below returns four (4) records.
```{r}
sample_n(flights, size=4)
```
Sampling can be performed with or without replacement,
and a weighting vector can be specified. 
See `help(sample_n)` for more details. 

## Function `sample_frac`

The `sample_frac` function also returns a random selection of records.
It returns a proportion of (the rows of) the data frame as determined by 
the `size` parameter.
The `flights` data frame contains `r nrow(flights)` records.
Three (3) is the nearest integer to 
`r nrow(flights)`/100000 and 
is the number of records returned by the command below.
```{r}
sample_frac(flights,size=1/100000)
```

## Function `slice`

The `slice` command retrieves specific rows of the input data frame
as indicated by the vector given in the second argument. 
For example, the following command returns the first four rows
of the data frame.
```{r}
slice(flights, 1:4)
```

The first and fourth rows of the data frames are returned from the command
```{r}
slice(flights, c(1,4))
```

The last row of the data frame is returned with the command
```{r}
slice(flights, n())
```

The last four (4) rows of the data frame are returned with the command
```{r}
slice(flights, (n()-3):n())
```

## Function `filter`

The `filter` command selects a subset of the rows of a data frame 
based on values of variables of the data frame. 
For example, all records where `month==12` _and_ `day==25` (flights on Dec 25) 
are requested with the following command. 
```{r}
flights.12.25 = filter(flights, month==12, day==25)
flights.12.25 
```
All conditions specified in the second and later arguments must be satisfied. 

To specify an _or_ condition use the pipe "|" symbol. 
The following command retrieves all rows where `month` is 
equal to either `11` _or_ `12`. 
```{r}
filter(flights, month==11 | month==12)
```

## Function `distinct`

__This command should be used with caution!__

In many cases I've seen it misused, which produces errors. 
If you do choose to use the `distinct` function please check and verify your
output to confirm that it is what you expect. 

The `distinct` function takes as input a data frame in the first argument, 
and a list of variables in the second and, possibly, later arguments. 
The function returns a data frame where the values of the listed variables
are unique between records. For example, the following command
returns `r nrow(distinct(flights,month))` rows, one for every 
unique value of  variable `month`.
```{r}
distinct(flights, month)
```

The next command
returns `r nrow(distinct(flights,month,day))` rows, one for every 
unique pair of values of variables `month` and `day`.
```{r}
flights.month.day = distinct(flights, month, day)
nrow(flights.month.day)
```

The record returned for each unique set of values of the input variables
is the _first_ record found with these values.
Unless the data frame is sorted it will be unclear which record is chosen. 
A more useful and general approach is to summarize the data, 
as described in Section _Summarizing Variables_ below. 

# Selecting and renaming variables 

## Function `select`

The `select` command returns a data frame with only the specified 
variables from the input data frame. 

```{r}
flights.dep = select(flights, month, day, starts_with('dep'))
str(flights.dep)
```
Notice the use of the `starts_with` function. 
There are additional options for specifying variables. 
See `help(select)` for details.

## Function `rename`

The `rename` function renames variables and returns all variables 
of the input data frame. 
```{r}
flights.departure = rename(flights.dep, 
                           departure.delay = dep_delay, 
                           departure.time  = dep_time
                           )
str(flights.departure)
```

# Modifying variables with the `mutate` and `transmute` functions

The `mutate` function alters variables of the input data frame
and returns the modified data frame. It can also create new variables. 
Note that the input data frame is not changed. 
For example, the following command changes the units of `dep_delay` to hours.
```{r}
mutate(flights.dep, dep_delay=dep_delay/60)
```

See also the `transmute` function which drops all variables which 
where not explicitly modified. 

# Summarizing variables with the `group_by` and `summarize` functions

Recall the descriptive statistics commands which return a single value 
given input of a numeric vector: 
`length`,  `sum`,  `mean`,  `median`, 
`sd` (standard deviation),  `var` (variance), 
`min` and `max`.
The `summarize` command in conjunction with these functions 
is used to summarize variables of a data frame. 
For example, the following command displays the minimum, maximum, 
mean and median values of  variable `dep_delay`.
```{r}
summarize(flights, 
          min=min(dep_delay,na.rm=TRUE),
          max=max(dep_delay,na.rm=TRUE),
          mean=mean(dep_delay,na.rm=TRUE),
          median=median(dep_delay,na.rm=TRUE))
```
Notice the parameter `na.rm=TRUE` which indicates that `NA` values 
should be removed prior to running these functions. 

Summarization of variables is often performed, not for the entire 
dataset, but for groups of rows as determined by a categorical variable. 
For example, with the command below we summarize the flights dataset for each group of rows as determined by the `month` variable. 

```{r}
summarize(group_by(flights,month), 
          min=min(dep_delay,na.rm=TRUE),
          max=max(dep_delay,na.rm=TRUE),
          mean=mean(dep_delay,na.rm=TRUE),
          median=median(dep_delay,na.rm=TRUE))
```

Multiple variables can be used to group rows of the dataset. 
For example, the following command produces summaries for each of 
`r  nrow(summarize(group_by(flights, month, day), 
          min=min(dep_delay,na.rm=TRUE),
          max=max(dep_delay,na.rm=TRUE),
          mean=mean(dep_delay,na.rm=TRUE),
          median=median(dep_delay,na.rm=TRUE)))` 
groups as determined by variables `month` and `day`.
```{r}
summarize(group_by(flights, month, day), 
          min=min(dep_delay,na.rm=TRUE),
          max=max(dep_delay,na.rm=TRUE),
          mean=mean(dep_delay,na.rm=TRUE),
          median=median(dep_delay,na.rm=TRUE))
```

# References

+ [dplyr: A Grammar of Data Manipulation](http://cran.r-project.org/web/packages/dplyr/index.html) (CRAN repository)
+ [Package dplyr](http://cran.r-project.org/web/packages/dplyr/dplyr.pdf)
  (documentation)
+ [Introduction to dplyr](http://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) (tutorial)
+ [Github: hadley/dplyr](https://github.com/hadley/dplyr)