---
title: "Analytics Without Borders - First R File"
author: "Arthur R. Markdown"
date: "March 31, 2017"
output: html_document
---

## Section 1 - Formatting Text

This forum explores all ways of _analytics work_, be it from corporate institutions, academia, or government organizations. 
The event will be a blend of learning opportunities including:

- Keynote talks
- Hands-on workshops in __R__, __Python__, and __Tableau__
- Research presentations
- Poster presentations

Notice that there is a blank line before and after the paragraph and before and after the list items.

## Section 2 - Code blocks

Many datasets are available in R. 
The first five rows of the `mtcars` dataset are:
```{r mtcars-head}
head(mtcars, 5)
```

Following is a summary of each of the variables in the `mtcars` dataset:
```{r mtcars-summary, echo=FALSE}
summary(mtcars)
```
Notice that in this second _code block_ that output was displayed, but that the command was not. 
This behavior is caused by the `echo=FALSE` parameter. 

## Conclusion

So that's cool. 
